TCP Rax: Deep Reinforcement Learning for TCP Congestion Control

### 2018-01-31 tex/paper.tex
Update paper

In a machine learning based congestion control method, each sender uses a set of observed environment conditions to determine which action he should take to maximize his reward in the future. The environment conditions can be any metrics that the sender can obtain from the network, for example the mean round-trip time of the last received packets, the packet loss rate etc. An action is a change to the sending rate: For example, if a sender perceives an increase in the loss rate it might be advisable to lower the sending rate. To modulate the sending rate one can  either
\begin{enumerate*}
\item change the sending rate itself at certain points in time or
\item maintain a congestion window that indicates the number of packets that can be in flight at a certain point in time
\end{enumerate*}. It seems preferable to maintain a congestion window, as one cannot be certain that the operating system (OS) can maintain a certain constant sending rate for each application on one machine, as the OS has to schedule different applications and share the network interfaces between them.

While congestion control seems to be a well suited problem for reinforcement learning, there are a couple of unique aspects and differences that distinguish it from most other reinforcement learning problems:  \begin{itemize}
\item \textbf{Delayed rewards}: In congestion control, rewards are always delayed by one round-trip time, while in most other areas where reinforcement learning is applied, if one performs an action, one can see the consequences of it action immediately. Also, due to the delay, by the time an action receives its reward, probably other actions have already been carried out in the meantime. Thus, contrary to most RL applications, actions and rewards are not synchronized. 
\item \textbf{Actions and rewards are not atomic}: If an action causes three packets to be sent, we consider these three packets being sent three \textit{partial actions}. In case these packets are transmitted correctly, the receiver will send back three acknowledgements. Each of these acknowledgements is a \textit{partial reward} that already unveils new information on the current state of the network and enables the sender to take an action based on this new information. However, only if all partial rewards have been received, the sender can assemble the overall reward of the corresponding action and can perform an update of the underlying reinforcement learning logic. 
\item \textbf{Variable number of actions and rewards}: If the congestion window is larger, more packets are sent and more acknowledgements are received, which means that the reinforcement learning algorithm gets more training data the larger the window is. This is problematic as it means that -- without proper adjustments -- the experience made at a larger congestion window will ``overwrite'' the experience made at a smaller window as with the larger window simply more training data arrives and it is thus going to be overemphasized. 
\end{itemize}

@@

\begin{figure}
\begin{minipage}{\columnwidth}
\includegraphics[width=\columnwidth]{{{/home/max/repos/remy/compute-0-4.local/figures/4_1_2018-01-25-20:17:15_3_2018-01-25-20:17:15.0}}}
\includegraphics[width=\columnwidth]{{{/home/max/repos/remy/compute-0-6.local/figures/4_2_2018-01-25-19:53:39_22_2018-01-25-19:53:39.0}}}
\caption{Throughput and loss rate when sharing a bottleneck link of 20\,Mbit/s with a two-way end-to-end delay of 50\,ms and a very small buffer of $\frac{1}{10}$ bandwidth delay product between 1 and 2 senders (from top to bottom). The time needed for convergence increases with more senders, which is caused by the fact that the more senders use the link, the fewer packets each sender receives per unit of time for training and because of the increasing complexity of the environment. It is interesting to note that in the case of two senders, both flows complement each other and quickly adapt to the sending rate of the other flow.}
\label{fig:demo}
\end{minipage}
\end{figure}

@@

The actor also aims at increasing the probability density function's entropy ($\frac{1}{2} \log \left( 2 \pi e \sigma^2 \right)$) to encourage exploration: Otherwise the actor could always perform similar actions but never actually take the action that yields the best performance. The actor network will aim to maximize the entropy and at the same time output the best possible action: If it realizes that the entropy is too large and bad actions are taken, it learns to lower the entropy. We designate the entropy as $H$ and introduce a parameter $\beta$ which specifies magnitude of the influence of the entropy. 
\begin{align}
l_{\text{a},t} =& -\log \left( \pi\left( a_t \given s_t; \theta \right)\right)\left( R_t - v_t \right)\\ 
&- \beta H\left( \pi\left( s_t; \theta \right)\right).
\end{align}

@@

\begin{description}
\item[$\textit{s}_\text{t}$] The state  describes the current ``congestion state'' defined by:
\begin{itemize}
\item the time between the last two packets that were received
\item the RTT of the last received packet
\item whether the last packet was lost, where $1$ designates that the last packet was lost while $0$ indicates that it was received correctly
\item the current congestion window (as a real number, as actions by output by the actor network can be very small and thus the congestion window is often only changed by minuscule amounts)%
\end{itemize}
Of each of these features a exponentially weighted moving average (ewma) with a factor $\alpha$ of $\frac{1}{8}$ and one with an $\alpha$ of $\frac{1}{256}$ (the concept of using ewmas with these specific values of $\alpha$ was first introduced in \textit{Remy} \cite{winstein_tcp_2013}), which makes a total of 8 features.  
Each time an acknowledgement is received, the state is updated and the actor network is asked for the next action.
\item[$\textit{a}_t$] Based on a given state and the history of previous states (because we use Gated Recurrent Units (GRUs) \cite{cho_learning_2014} and thus can also consider previous states), the actor network returns an action $a_t$, which is a real number that stands for the change of the congestion window.

@@

In case of congestion control, the loss function $l_{\text{v},t}$ of the critic network is actually the sum of the squares of the difference for each of the expected long-term averages and the empirically found averages for each reward metric:
\begin{align}
l_{\text{v},t} =& \left(R_{\text{received},t} - V_\text{received}(s_t; \theta)\right)^2 \\
&+\left(R_{\text{sent},t} - V_\text{sent}(s_t; \theta)\right)^2 \\
&+\left(R_{\text{duration},t} - V_\text{duration}(s_t; \theta)\right)^2
\end{align}

@@

\begin{align}
l_{\text{a},t} =& -\log \left( \pi \left( a_t \given s_t ; \theta_\text{a} \right) \right)\\
& \left( U_{\text{measured},t} - U_{\text{expected},t} \right)\\ 
&- \beta H\left( \pi\left( s_t; \theta_a \right)\right)
\end{align}
where $U$ can be any utility function defined based on some of the (long term averages of the) reward metrics. Thus, in simple terms, the actor network considers if an action improved the actual experienced utility compared to the expected one and adjusts the neural network accordingly.

>>>>

In a machine learning based congestion control method, each sender uses a set of observed environment conditions to determine which action it should take to maximize his reward in the future. The environment conditions can be any metrics that the sender can obtain from the network, for example the mean round-trip time of the last received packets, the packet loss rate etc. An action is a change to the sending rate: For example, if a sender perceives an increase in the loss rate it might be advisable to lower the sending rate. To modulate the sending rate one can  either
\begin{enumerate*}
\item change the sending rate itself at certain points in time or
\item maintain a congestion window that indicates the number of packets that can be in flight at a certain point in time
\end{enumerate*}. It seems preferable to maintain a congestion window, as one cannot be certain that the operating system (OS) can maintain a certain constant sending rate for each application on one machine, as the OS has to schedule different applications and share the network interfaces between them.

While congestion control seems to be a well-suited problem for reinforcement learning, there are a couple of unique aspects and differences that distinguish it from most other reinforcement learning problems:  \begin{itemize}
\item \textbf{Delayed rewards}: In congestion control, rewards are always delayed by one round-trip time, while in most other areas where reinforcement learning is applied, if one performs an action, one can see the consequences of it action immediately. Also, due to the delay, by the time an action receives its reward, probably other actions have already been carried out in the meantime. Thus, contrary to most RL applications, actions and rewards are not synchronized. 
\item \textbf{Actions and rewards are not atomic}: If an action causes three packets to be sent, we consider these three packets being sent three \textit{partial actions}. In case these packets are transmitted correctly, the receiver will send back three acknowledgements. Each of these acknowledgements is a \textit{partial reward} that already unveils new information on the current state of the network and enables the sender to take an action based on this new information. However, only if all partial rewards have been received, the sender can assemble the overall reward of the corresponding action and can perform an update of the underlying reinforcement learning logic. 
\item \textbf{Variable number of actions and rewards}: If the congestion window is larger, more packets are sent and more acknowledgements are received, which means that the reinforcement learning algorithm gets more training data the larger the window is. This is problematic as it means that -- without proper adjustments -- the experience made at a larger congestion window will ``overwrite'' the experience made at a smaller window as with the larger window simply more training data arrives and it is thus going to be overemphasized. 
\end{itemize}

@@

\begin{figure}
\begin{minipage}{\columnwidth}
\subfloat[One sender]{\includegraphics[width=\columnwidth]{{{/home/max/repos/remy/compute-0-4.local/figures/4_1_2018-01-25-20:17:15_3_2018-01-25-20:17:15.0}}}}{}
\subfloat[Two senders]{\includegraphics[width=\columnwidth]{{{/home/max/repos/remy/compute-0-6.local/figures/4_2_2018-01-25-19:53:39_22_2018-01-25-19:53:39.0}}}}{}
\caption{Throughput and loss rate when sharing a bottleneck link of 20\,Mbit/s with a two-way end-to-end delay of 50\,ms and a very small buffer of $\frac{1}{10}$ bandwidth delay product between 1 and 2 senders (from top to bottom). The time needed for convergence increases with more senders, which is caused by the fact that the more senders use the link, the fewer packets each sender receives per unit of time for training and because of the increasing complexity of the environment. It is interesting to note that in the case of two senders, both flows complement each other and quickly adapt to the sending rate of the other flow.}
\label{fig:demo}
\end{minipage}
\end{figure}

@@

The actor also aims at increasing the probability density function's entropy ($\frac{1}{2} \log \left( 2 \pi e \sigma^2 \right)$) to encourage exploration: Otherwise the actor could always perform similar actions but never actually take the action that yields the best performance. The actor network will aim to maximize the entropy and at the same time output the best possible action: If it realizes that the entropy is too large and bad actions are taken, it learns to lower the entropy. We designate the entropy as $H$ and introduce a parameter $\beta$ which specifies magnitude of the influence of the entropy. The loss function of the actor network is defined as follows \cite{mnih_asynchronous_2016}:
\begin{align}
\begin{split}
l_{\text{a},t} =& -\log \left( \pi\left( a_t \given s_t; \theta \right)\right)\left( R_t - v_t \right)\\ 
&- \beta H\left( \pi\left( s_t; \theta \right)\right).
\end{split}
\end{align}

@@

\begin{description}
\item[$\textit{s}_\text{t}$] The state  describes the current ``congestion state'' defined by:
\begin{itemize}
\item the time between the last two packets that were received
\item the RTT of the last received packet
\item whether the last packet was lost, where $1$ denotes that the last packet was lost while $0$ indicates that it was received correctly
\item the current congestion window (as a real number, as actions by output by the actor network can be very small and thus the congestion window is often only changed by minuscule amounts)%
\end{itemize}
Of each of these features a exponentially weighted moving average (ewma) with a factor $\alpha$ of $\frac{1}{8}$ and one with an $\alpha$ of $\frac{1}{256}$ (the concept of using ewmas with these specific values of $\alpha$ was first introduced in \textit{Remy} \cite{winstein_tcp_2013}), which makes a total of 8 features.  
Each time an acknowledgement is received, the state is updated and the actor network is asked for the next action.
\item[$\textit{a}_t$] Based on a given state and the history of previous states (because we use Gated Recurrent Units (GRUs) \cite{cho_learning_2014} and thus can also consider previous states), the actor network returns an action $a_t$, which is a real number that stands for the change of the congestion window.

@@

In case of congestion control, the loss function $l_{\text{v},t}$ of the critic network is actually the sum of the squares of the difference for each of the expected long-term averages and the empirically found averages for each reward metric:
\begin{align}
\begin{split}
l_{\text{v},t} =& \left(R_{\text{received},t} - V_\text{received}(s_t; \theta)\right)^2 \\
&+\left(R_{\text{sent},t} - V_\text{sent}(s_t; \theta)\right)^2 \\
&+\left(R_{\text{duration},t} - V_\text{duration}(s_t; \theta)\right)^2
\end{split}
\end{align}

@@

\begin{align}
\begin{split}
l_{\text{a},t} =& -\log \left( \pi \left( a_t \given s_t ; \theta_\text{a} \right) \right)\left( U_{\text{measured},t} - U_{\text{expected},t} \right)\\ 
&- \beta H\left( \pi\left( s_t; \theta_a \right)\right)
\end{split}
\end{align}
where $U$ can be any utility function defined based on some of the (long term averages of the) reward metrics. Thus, in simple terms, the actor network considers if an action improved the actual experienced utility compared to the expected one and adjusts the neural network accordingly.
